---
title: "how_to_code_in_r.Rmd"
author: "Sophia Ungar"
date: "2024-01-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


I'm not going to go into why use R, except that it seems to be primarily for statistics. I'm not the best person to ask about R vs other stats languages vs Python. I also want to make this a useful reference guide (mostly for myself), but not a complete teacher.


## Other places to learn things
Suggested by Josh Davidson:
[R for Data Science](https://r4ds.hadley.nz/data-visualize)


## Basic Formatting
RMarkdown combines R code in chunks like this one: 
```{r}

```
The three ticks and the {r} as seen above designate the things inside as things to be run. You can leave comments that start with a pound sign, #, that won't be evaluated.

Outside of the chunks (sorry about the word, not my choice), text is formatted with markdown. 
About Markdown: https://www.markdownguide.org/

RMarkdown files can be knit to see the final form. Click on the ball of yarn/ Knit button above this. Or click the drop down arrow to choose whether to end up with a pdf, HTML, or Word document.


## Basic Functions

```{r}
# Arithmetic
2+3
2-4
8*9
5/10
4**10

# Variable Assignment
a <- 2
b = 4

# Print
b
5

```


## Extra code we need
If you haven't installed the packages first you will need to run  install.packages("<package>") in the terminal below or go to packages in the bottom right.
```{r}
library(tidyverse)
library(ggplot2) # for the visualizations
```


## Read in data
```{r}
#data <- read.csv(<filename>.csv) # getting data from a csv
library(palmerpenguins)
data <- penguins
```
We want to pull in data so the computer can look at it. I've commented out one way to get it from a csv. To use as an example, the palmerpenguins library has the penguins tibble. 

One thing to think about is what kind of structure we want to put the data in. In R there are tables, data frames, tibbles, and more. Some functions work on all of them, but some only work on certain types.
TODO: add recommendation on which structure to choose



## Clean Up 
Get a specific column with dataset_name$column_name.
```{r}
flipper <- data$flipper_length_mm

```


Isolate or Remove columns
```{r}
# same data but only the flipper and bill length columns
flipper_bill <- subset(data, select=c(flipper_length_mm, bill_length_mm))

# same data but without the island and year columns
no_island_year <- subset(data, select=-c(island, year))

```


TODO: add Remove rows with missing data
TODO: add find and remove outliers
TODO: take the log of everything?


## Summarize
In R there are many ways to do the same thing, in part because it is open source so many different people write the packages that provide a lot of the functions. Below is one way to start looking at the data.

```{r}
summary(data)

head(data) # see the first rows. 6 rows by default
```


## Basic Stats Functions
```{r}
mean( data$flipper_length_mm, na.rm=TRUE) # mean, ignoring missing values
sd( flipper, na.rm=TRUE) # standard deviation, ignoring missing values

```
TODO: add confidence intervals

t test
```{r}
# test only works to compare two levels
# so here we just get Chinstrap and Adelie
chinstrap_adelie <- filter( data, species=="Chinstrap" | species=="Adelie")

# bill length depends on species
t.test(bill_length_mm ~ species, chinstrap_adelie)

```

linear regression
```{r}
lm(bill_length_mm ~ body_mass_g, data = data)

```


ANOVA
```{r}
summary( aov(bill_length_mm ~ year, data))

```


## Visualize

This example shows flipper length by bill length. The color is the sex of the bird.
```{r}
ggplot(data, 
       aes(flipper_length_mm, bill_length_mm, color=sex)) +
    geom_point()

```


## Simulate
Simulate from the normal and uniform distributions.
```{r}
rnorm(n = 10, mean = 100, sd = 20)
runif(n = 5, min = 0, max = 1)

```

Bootstrapping
I think this was described as a more accurate way to accept or reject the null hypothesis.
```{r}
library(boot)
set.seed(23) # for replicability

our_function <- function(formula, data, indices) {
  fit <- lm(formula, data[indices,]) # linear regression
  
  return( summary(fit)$r.square)
}

# bootstrap 10,000 times
# see how much body mass and bill depth explains bill length?
boot_obj <- boot(data = data, statistic = our_function, R=10000, formula = bill_length_mm ~ body_mass_g + bill_depth_mm)

boot_obj
plot(boot_obj)
boot.ci(boot_obj, type="bca")

```
Note that our_function takes in a formula as an argument.

adapted from: https://www.datacamp.com/doc/r/bootstrapping
